```{r hw2_setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

Stats is pooopy

nice

# Homework 2 {-}

<center>
**Due Friday, February 18 at 9:00am CST on [Moodle](https://moodle.macalester.edu/mod/assign/view.php?id=36784)**
</center>

**Deliverables:** Please use [this template](template_rmds/hw2.Rmd) to knit an HTML document. Convert this HTML document to a PDF by opening the HTML document in your web browser. *Print* the document (Ctrl/Cmd-P) and change the destination to "Save as PDF". Submit this one PDF to Moodle.

Alternatively, you may knit your Rmd directly to PDF if you have LaTeX installed.

<br><br><br>

## Project Work {-}

### Instructions {-} 

**Goal:** Begin an analysis of your dataset to answer your **regression** research question.

<br>

**Collaboration:** Form a team (2-3 members) for the project and this part can be done as a team. Only one team member should submit a Project Work section. Make sure you include the full names of all of the members in your write up. 

<br>

**Data cleaning:** If your dataset requires any cleaning (e.g., merging datasets, creation of new variables), first consult the [R Resources page](r-resources.html) to see if your questions are answered there. If not, post on the #rcode-questions channel in our Slack workspace to ask for help. *Please ask for help early and regularly* to avoid stressful workloads.

<br>

### Project Work
1. Finalize the data set you'd like to work with
- COVID data from Our World in Data

2. Make sure that you can read it into R
- YES

3. Made some initial progress on visualizing and cleaning the data
  + Choose one quantitative outcome variable for regression
  
    Number of confirmed cases, `total_cases`
  
  + Focus on a subset of about 10-20 predictor variables for now
  
    -   `location`
    -   `continent`
    -   `date`
    -   `population_density` 
    -   `total_deaths`
    -   `reproduction_rate`
    -   `people_vaccinated_per_hundred`
    -   `people_fully_vaccinated_per_hundred`
    -   `total_boosters_per_hundred`
    -   `tests_per_case`
    -   `handwashing_facilities`
    -   `human_development_index`
    
  + Focus on a subset of rows that you can easily load into R
  
    -   We will filter the data to be about a specific set of dates  
  
  + Look for missing data & outliers
  
    -   A lot of rows with earlier dates in time have empty values
    -   This makes sense because the pandemic had maybe not reached those places yet nor had efforts started to collect data there
    
  + Understand what the values of the variables mean
    
    -   They are explained in the [https://www.kaggle.com/georgesaavedra/covid19-dataset](Kaggle page about the data)
  
  + Get a sense for possible relationships that may exist in the data
  
    -   Coming into this project, we already suspect that that there will be a strong negative correlation between vaccination metrics and the number of total cases that are reported in any location
  

### Required Analyses {-}

1. **Initial investigation: ignoring nonlinearity (for now)**
    a. Use ordinary least squares (OLS) by using the `lm` engine and LASSO (`glmnet` engine) to build  a series of initial regression models for your quantitative outcome as a function of the predictors of interest. (As part of data cleaning, exclude any variables that you don't want to consider as predictors.)
        - You'll need two model specifications, `lm_spec` and `lm_lasso_spec` (you'll need to tune this one).
        
    b. For each set of variables, you'll need a `recipe` with the `formula`, `data`, and pre-processing steps
        - You may want to have steps in your recipe that remove variables with near zero variance (`step_nzv()`), remove variables that are highly correlated with other variables (`step_corr()`), normalize all quantitative predictors (`step_normalize(all_numeric_predictors())`) and add indicator variables for any categorical variables (`step_dummy(all_nominal_predictors())`).
        - These models should not include any transformations to deal with nonlinearity. You'll explore this in the next investigation.
        
    c. Estimate the test performance of the models using CV. Report and interpret (with units) the CV metric estimates along with a measure of uncertainty in the estimate (`std_error` is readily available when you used `collect_metrics(summarize=TRUE)`).
        - Compare estimated test performance across the models. Which models(s) might you prefer?
        
    d. Use residual plots to evaluate whether some quantitative predictors might be better modeled with nonlinear relationships.
    
    e. Which variables do you think are the most important predictors of your quantitative outcome? Justify your answer. Do the methods you've applied reach consensus on which variables are most important? What insights are expected? Surprising?
        - Note that if some (but not all) of the indicator terms for a categorical predictor are selected in the final models, the whole predictor should be treated as selected.

<br>

#### Your Work {-}

a & b.

```{r}
# library statements 
library(dplyr)
library(readr)
library(broom)
library(ggplot2)
library(tidymodels) 
tidymodels_prefer() # Resolves conflicts, prefers tidymodel functions

# read in data
covid <- read_csv("owid-covid-data.csv")
```

```{r}
# data cleaning
covid <- covid %>% 
  select(total_cases, new_cases, location, continent, date, population_density, total_deaths, reproduction_rate, people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred, total_boosters_per_hundred, tests_per_case, handwashing_facilities, human_development_index)
```

- The date range is May 1st 2021 onwards. 
- This is when vaccinations started to become widespread (at least in the United States).
- We will start by looking at worldwide trends, and apply a `filter` if we want to look at a specific country or region.

```{r}
# creation of cv folds
set.seed(123)
covid_cv10 <- vfold_cv(covid, v = 10)
```

```{r}
# model spec
# OLS
lm_spec <-
    linear_reg() %>% 
    set_engine(engine = 'lm') %>% 
    set_mode('regression')

# LASSO
lm_lasso_spec <- 
  linear_reg() %>%
  set_args(mixture = 1, penalty = tune()) %>% 
  set_engine(engine = 'glmnet') %>% 
  set_mode('regression') 

```

```{r}
# recipes & workflows
# specify Recipe (if you have preprocessing steps)
covid_recipe <- recipe(total_cases ~ reproduction_rate + people_vaccinated_per_hundred, data = covid) %>%
  step_nzv(all_predictors()) %>%
  step_novel(all_nominal_predictors()) %>% 
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_naomit(all_predictors(), skip = TRUE)

covid_wf <- workflow() %>%
  add_recipe(covid_recipe) %>%
  add_model(lm_spec)

covid_wf_lasso <- workflow() %>%
  add_recipe(covid_recipe) %>%
  add_model(lm_lasso_spec)
```

```{r}
# fit & tune models
mod_lm_fit <- fit(covid_wf, data = covid)
mod_lm_fit %>% tidy()

mod_lasso_fit <- fit(covid_wf_lasso, data = covid)

plot(mod_lasso_fit %>% extract_fit_parsnip() %>% pluck('fit'), # way to get the original glmnet output
     xvar = "lambda") # glmnet fits the model with a variety of lambda penalty values

# tune model (try a variety of lambda penalty values)
penalty_grid <- grid_regular(penalty(range = c(-5,3)), levels = 30)




```

c.

```{r}

tune_res <- tune_grid( # new function for tuning parameters
  covid_wf_lasso, # workflow
  resamples = covid_cv10, # cv folds
  metrics = metric_set(rmse, mae),
  grid = penalty_grid # penalty grid defined above
)

#  calculate/collect CV metrics
collect_metrics(tune_res) %>%
  filter(.metric == 'rmse') %>% # or choose mae
  select(penalty, rmse = mean) 

# choose penalty value based on lowest mae or rmse
best_penalty <- select_best(tune_res, metric = 'rmse') 


```

 
d.

```{r}
# visual residuals
```

e.

<br>

2. **Summarize investigations**
    - Decide on an overall best model based on your investigations so far. To do this, make clear your analysis goals. Predictive accuracy? Interpretability? A combination of both?



<br>

3. **Societal impact**
    - Are there any harms that may come from your analyses and/or how the data were collected?
    




<br><br><br>
