```{r hw2_setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```



# Homework 2 {-}

<center>
**Due Friday, February 18 at 9:00am CST on [Moodle](https://moodle.macalester.edu/mod/assign/view.php?id=36784)**
</center>

**Deliverables:** Please use [this template](template_rmds/hw2.Rmd) to knit an HTML document. Convert this HTML document to a PDF by opening the HTML document in your web browser. *Print* the document (Ctrl/Cmd-P) and change the destination to "Save as PDF". Submit this one PDF to Moodle.

Alternatively, you may knit your Rmd directly to PDF if you have LaTeX installed.

<br><br><br>

## Project Work {-}

### Instructions {-} 

**Goal:** Begin an analysis of your dataset to answer your **regression** research question.

<br>

**Collaboration:** Form a team (2-3 members) for the project and this part can be done as a team. Only one team member should submit a Project Work section. Make sure you include the full names of all of the members in your write up. 

<br>

**Data cleaning:** If your dataset requires any cleaning (e.g., merging datasets, creation of new variables), first consult the [R Resources page](r-resources.html) to see if your questions are answered there. If not, post on the #rcode-questions channel in our Slack workspace to ask for help. *Please ask for help early and regularly* to avoid stressful workloads.

<br>

### Project Work
1. Finalize the data set you'd like to work with
- COVID data from Our World in Data

2. Make sure that you can read it into R
- YES

3. Made some initial progress on visualizing and cleaning the data
  + Choose one quantitative outcome variable for regression
  
    Number of confirmed cases, `total_cases`
  
  + Focus on a subset of about 10-20 predictor variables for now
  
    -   `date`
    -   `total_deaths`
    -   `reproduction_rate`
    -   `people_vaccinated_per_hundred`
    -   `people_fully_vaccinated_per_hundred`
    -   `total_boosters_per_hundred`
    -   `tests_per_case`
    -   `new_cases`
    
  + Focus on a subset of rows that you can easily load into R
  
    -   We're focusing on cases within the USA. 
  
  + Look for missing data & outliers
  
    -   A lot of rows with earlier dates in time have empty values
    -   This makes sense because the pandemic had maybe not reached those places yet nor had efforts started to collect data there
    
  + Understand what the values of the variables mean
    
    -   They are explained in the [https://www.kaggle.com/georgesaavedra/covid19-dataset](Kaggle page about the data)
  
  + Get a sense for possible relationships that may exist in the data
  
    -   Coming into this project, we already suspect that that there will be a strong negative correlation between vaccination metrics and the number of total cases that are reported in any location.
  

### Required Analyses {-}

1. **Initial investigation: ignoring nonlinearity (for now)**
    a. Use ordinary least squares (OLS) by using the `lm` engine and LASSO (`glmnet` engine) to build  a series of initial regression models for your quantitative outcome as a function of the predictors of interest. (As part of data cleaning, exclude any variables that you don't want to consider as predictors.)
        - You'll need two model specifications, `lm_spec` and `lm_lasso_spec` (you'll need to tune this one).
        
    b. For each set of variables, you'll need a `recipe` with the `formula`, `data`, and pre-processing steps
        - You may want to have steps in your recipe that remove variables with near zero variance (`step_nzv()`), remove variables that are highly correlated with other variables (`step_corr()`), normalize all quantitative predictors (`step_normalize(all_numeric_predictors())`) and add indicator variables for any categorical variables (`step_dummy(all_nominal_predictors())`).
        - These models should not include any transformations to deal with nonlinearity. You'll explore this in the next investigation.
        
    c. Estimate the test performance of the models using CV. Report and interpret (with units) the CV metric estimates along with a measure of uncertainty in the estimate (`std_error` is readily available when you used `collect_metrics(summarize=TRUE)`).
        - Compare estimated test performance across the models. Which models(s) might you prefer?
        
    d. Use residual plots to evaluate whether some quantitative predictors might be better modeled with nonlinear relationships.
    
    e. Which variables do you think are the most important predictors of your quantitative outcome? Justify your answer. Do the methods you've applied reach consensus on which variables are most important? What insights are expected? Surprising?
        - Note that if some (but not all) of the indicator terms for a categorical predictor are selected in the final models, the whole predictor should be treated as selected.

<br>

#### Your Work {-}

a & b.

```{r}
# library statements 
library(dplyr)
library(readr)
library(broom)
library(ggplot2)
library(tidymodels)
library(lubridate)
library(gridExtra)
tidymodels_prefer() # Resolves conflicts, prefers tidymodel functions

# read in data
covid <- read_csv("owid-covid-data.csv")
```

```{r}
# data cleaning
covid <- covid %>% 
  filter(location == 'United States') %>%
  select(total_cases, new_cases, date, total_deaths, reproduction_rate, people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred, total_boosters_per_hundred, tests_per_case, icu_patients, positive_rate) %>%
  filter(date >= as.Date("2021-08-13"))%>%
  na.omit(covid)
```

- We're considering COVID cases occurring after the first recorded booster vaccines to best capture today's landscape.  
- This is when vaccinations started to become widespread (at least in the United States).
- We will start by looking at worldwide trends, and apply a `filter` if we want to look at a specific country or region.

```{r}
# creation of cv folds
set.seed(123)
covid_cv10 <- vfold_cv(covid, v = 10)
```

```{r}
# model spec
# OLS
lm_spec <-
    linear_reg() %>% 
    set_engine(engine = 'lm') %>% 
    set_mode('regression')

mod1 <- fit(lm_spec,
            new_cases ~ ., 
            data = covid)

mod2 <- fit(lm_spec,
            new_cases ~ reproduction_rate, 
            data = covid)

mod3 <- fit(lm_spec,
            new_cases ~ reproduction_rate+people_vaccinated_per_hundred, 
            data = covid)

mod4 <- fit(lm_spec,
            new_cases ~ reproduction_rate+people_vaccinated_per_hundred+people_fully_vaccinated_per_hundred, 
            data = covid)

mod5 <- fit(lm_spec,
            new_cases ~ reproduction_rate+people_vaccinated_per_hundred+people_fully_vaccinated_per_hundred+total_boosters_per_hundred, 
            data = covid)


covid_rec_ols <- recipe(new_cases ~ .,  data = covid) %>%
  step_normalize(all_numeric_predictors()) 

#mod1
cov_wf_ols <- workflow() %>%
  add_recipe(covid_rec_ols) %>%
  add_model(lm_spec)%>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_nzv(all_predictors())



mod1 <- fit(cov_wf_ols,
            data = covid) 

#mod2
cov_wf_ols2 <- workflow() %>%
  add_formula(new_cases ~ reproduction_rate)%>%
  add_model(lm_spec)%>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_nzv(all_predictors())

cov_wf_ols3 <- workflow() %>%
  add_formula(new_cases ~ reproduction_rate+people_vaccinated_per_hundred)%>%
  add_model(lm_spec)%>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_nzv(all_predictors())

cov_wf_ols4 <- workflow() %>%
  add_formula(new_cases ~ reproduction_rate+people_vaccinated_per_hundred+people_fully_vaccinated_per_hundred)%>%
  add_model(lm_spec)%>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_nzv(all_predictors())

cov_wf_ols5 <- workflow() %>%
  add_formula(new_cases ~ reproduction_rate+people_vaccinated_per_hundred+people_fully_vaccinated_per_hundred+total_boosters_per_hundred)%>%
  add_model(lm_spec)%>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_nzv(all_predictors())

```

```{r}
mod1 %>% #comment here
  tidy() %>% #
  slice(-1) %>% # 
  mutate(lower = estimate - 1.96*std.error, upper = estimate + 1.96*std.error) %>% #
  ggplot() + # 
    geom_vline(xintercept=0, linetype=4) + # 
    geom_point(aes(x=estimate, y=term)) + # 
    geom_segment(aes(y=term, yend=term, x=lower, xend=upper), arrow = arrow(angle=90, ends='both', length = unit(0.1, 'cm'))) + # 
    labs(x = 'Coefficient estimate (95% CI)', y = 'Feature') +  #
    theme_classic() #

mod1 %>%
  tidy()
```
```{r}
mod1_output <- mod1 %>% 
    predict(new_data = covid) %>%
    bind_cols(covid) %>%
    mutate(resid = new_cases - .pred)

mod2_output <- mod2 %>% 
    predict(new_data = covid) %>%
    bind_cols(covid) %>%
    mutate(resid = new_cases - .pred)

mod3_output <- mod3 %>% 
    predict(new_data = covid) %>%
    bind_cols(covid) %>%
    mutate(resid = new_cases - .pred)

mod4_output <- mod4 %>% 
    predict(new_data = covid) %>%
    bind_cols(covid) %>%
    mutate(resid = new_cases - .pred)

mod5_output <- mod5 %>% 
    predict(new_data = covid) %>%
    bind_cols(covid) %>%
    mutate(resid = new_cases - .pred)
```

```{r}
mod1_output %>%
    mae(truth = new_cases, estimate = .pred)

mod2_output %>%
    mae(truth = new_cases, estimate = .pred)

mod3_output %>%
    mae(truth = new_cases, estimate = .pred)

mod4_output %>%
    mae(truth = new_cases, estimate = .pred)

mod5_output %>%
    mae(truth = new_cases, estimate = .pred)
```

```{r}
#mod1

ggplot(mod1_output, aes(y=resid, x=total_cases)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod1_output, aes(y=resid, x=total_deaths)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod1_output, aes(y=resid, x=reproduction_rate)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod1_output, aes(y=resid, x=people_vaccinated_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod1_output, aes(y=resid, x=people_fully_vaccinated_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod1_output, aes(y=resid, x=total_boosters_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod1_output, aes(y=resid, x=tests_per_case)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod1_output, aes(y=resid, x=date)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()
```

```{r}
#mod2
ggplot(mod2_output, aes(y=resid, x=reproduction_rate)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

```

```{r}

#mod3
ggplot(mod3_output, aes(y=resid, x=reproduction_rate)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod3_output, aes(y=resid, x=people_vaccinated_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

```
```{r}
#mod4

ggplot(mod4_output, aes(y=resid, x=reproduction_rate)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod4_output, aes(y=resid, x=people_vaccinated_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod4_output, aes(y=resid, x=people_fully_vaccinated_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()


```
```{r}
#mod5
ggplot(mod5_output, aes(y=resid, x=reproduction_rate)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod5_output, aes(y=resid, x=people_vaccinated_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod5_output, aes(y=resid, x=people_fully_vaccinated_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

ggplot(mod5_output, aes(y=resid, x=total_boosters_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

```

From the residual plots, it seems that `people_fully_vaccinated_per_hundred	`, `people_vaccinated_per_hundred` and	
`total_boosters_per_hundred` are variables that look like they would be better modeled with nonlinear relationships.  

```{r}
mod1_cv <- fit_resamples(cov_wf_ols,
  resamples = covid_cv10, 
  metrics = metric_set(rmse, rsq, mae))
  
  
mod2_cv <- fit_resamples(cov_wf_ols2,
  resamples = covid_cv10, 
  metrics = metric_set(rmse, rsq, mae))
  
  
mod3_cv <- fit_resamples(cov_wf_ols3,
  resamples = covid_cv10, 
  metrics = metric_set(rmse, rsq, mae))
  
  
mod4_cv <- fit_resamples(cov_wf_ols4,
  resamples = covid_cv10, 
  metrics = metric_set(rmse, rsq, mae))
  
  
mod5_cv <- fit_resamples(cov_wf_ols5,
  resamples = covid_cv10, 
  metrics = metric_set(rmse, rsq, mae))

```

```{r}

mod1_cv %>% unnest(.metrics) %>%
  filter(.metric == 'rmse') %>%
  summarize(RMSE_CV = mean(.estimate))

mod2_cv %>% unnest(.metrics) %>%
  filter(.metric == 'rmse') %>%
  summarize(RMSE_CV = mean(.estimate))

mod3_cv %>% unnest(.metrics) %>%
  filter(.metric == 'rmse') %>%
  summarize(RMSE_CV = mean(.estimate))

mod4_cv %>% unnest(.metrics) %>%
  filter(.metric == 'rmse') %>%
  summarize(RMSE_CV = mean(.estimate))

mod5_cv %>% unnest(.metrics) %>%
    filter(.metric == 'rmse') %>%
    summarize(RMSE_CV = mean(.estimate))

mod1_cv %>% collect_metrics()

mod2_cv %>% collect_metrics()

mod3_cv %>% collect_metrics()

mod4_cv %>% collect_metrics()

mod5_cv %>% collect_metrics()


```


```{r}
# LASSO
lm_lasso_spec <- 
  linear_reg() %>%
  set_args(mixture = 1, penalty = tune()) %>% 
  set_engine(engine = 'glmnet') %>% 
  set_mode('regression') 

```

```{r}
# recipes & workflows
# specify Recipe (if you have preprocessing steps)

covid <- covid %>% select(-date)

covid_recipe <- recipe(new_cases ~ ., data = covid) %>%
  step_nzv(all_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_naomit(all_predictors()) #skip = TRUE


covid_wf <- workflow() %>%
  add_recipe(covid_recipe) %>%
  add_model(lm_spec)

covid_wf_lasso <- workflow() %>%
  add_recipe(covid_recipe) %>%
  add_model(lm_lasso_spec)
```

```{r}
# fit & tune models

mod_lasso_fit <- fit(covid_wf_lasso, data = covid)

plot(mod_lasso_fit %>% extract_fit_parsnip() %>% pluck('fit'), # way to get the original glmnet output
     xvar = "lambda") # glmnet fits the model with a variety of lambda penalty values

# tune model (try a variety of lambda penalty values)
penalty_grid <- grid_regular(penalty(range = c(-5,3)), levels = 30)
```

c.

```{r}

tune_res <- tune_grid( # new function for tuning parameters
  covid_wf_lasso, # workflow
  resamples = covid_cv10, # cv folds
  metrics = metric_set(rmse, mae),
  grid = penalty_grid # penalty grid defined above
)

#  calculate/collect CV metrics
collect_metrics(tune_res) %>%
  filter(.metric == 'mae') %>% # or choose mae
  select(penalty, mae = mean, std_err) 

# best penalty
best_penalty <- select_best(tune_res, metric = 'mae') # choose penalty value based on lowest mae or rmse
best_se_penalty <- select_by_one_std_err(tune_res, metric = 'mae', desc(penalty)) # choose penalty value based on the largest penalty within 1 se of the lowest CV MAE
```

```{r}
#Fit Final Model
final_wf <- finalize_workflow(covid_wf_lasso, best_penalty) # incorporates penalty value to workflow
final_wf_se <- finalize_workflow(covid_wf_lasso, best_se_penalty) # incorporates penalty value to workflow

final_fit <- fit(final_wf, data = covid)
final_fit_se <- fit(final_wf_se, data = covid)

tidy(final_fit)
tidy(final_fit_se)
```

 
d.

```{r}
# visual residuals

tune_res %>% collect_metrics() %>% filter(penalty == (best_se_penalty %>% pull(penalty)))
```

### Residual Plot of LASSO Model Prediction

```{r}
lasso_mod_out <- final_fit_se %>%
    predict(new_data = covid) %>%
    bind_cols(covid) %>%
    mutate(resid = new_cases - .pred)

lasso_mod_out %>% 
  ggplot(aes(x = .pred, y = resid)) + 
  geom_point() +
  geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0) + 
  theme_classic()
```

### Residuals from LASSO Model

Our group determined from homework 1 that the LASSO model was the best model for our data so far, because it had the lowest MAE compared to the other models we tried.

```{r}
pl1 <- ggplot(lasso_mod_out, aes(y=resid, x=total_cases)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

pl2 <- ggplot(lasso_mod_out, aes(y=resid, x=total_deaths)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

pl3 <- ggplot(lasso_mod_out, aes(y=resid, x=reproduction_rate)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

pl4<- ggplot(lasso_mod_out, aes(y=resid, x=people_vaccinated_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

pl5 <- ggplot(lasso_mod_out, aes(y=resid, x=people_fully_vaccinated_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

pl6 <- ggplot(lasso_mod_out, aes(y=resid, x=total_boosters_per_hundred)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

pl7 <- ggplot(lasso_mod_out, aes(y=resid, x=tests_per_case)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

pl8 <- ggplot(lasso_mod_out, aes(y=resid, x=icu_patients)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

pl9 <- ggplot(lasso_mod_out, aes(y=resid, x=positive_rate)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") + 
    theme_classic()

pl1
pl2
pl3
pl4
pl5
pl6
pl7
pl8
pl9
```

From the residual plots of the LASSO model above, it looks like we can try to model all of these predictors with nonlinear relationships. This is because the plots show heteroskedasticity. 

e.

We can look at the LASSO regression results to determine variable importance. LASSO is a shrinking algorithm which will eliminate weak predictors from the model by setting their coefficients to 0. From the estimate values of the coefficients, we can see that `total_deaths`, `people_vaccinated_per_hundred` and `total_boosters_per_hundred` have been shrunk to 0 and are not considered important variables in the model. 

However, the magnitude of the other coefficients produced by the LASSO algorithm are quite large, so this indicates that they are all important variables.


<br>

2. **Non linearity**
```{r}
# GAMS

gam_spec <- 
  gen_additive_mod() %>%
  set_engine(engine = 'mgcv') %>%
  set_mode('regression') 

gam_mod <- fit(gam_spec,
    new_cases ~ s(total_cases) + s(total_deaths) + s(total_deaths) + s(reproduction_rate) + s(people_vaccinated_per_hundred) + s(people_fully_vaccinated_per_hundred) + s(tests_per_case) + s(icu_patients) + s(positive_rate) + s(total_boosters_per_hundred),
    data = covid
)

# Diagnostics: Check to see if the number of knots is large enough (if p-value is low, increase number of knots)
par(mfrow=c(2,2))
gam_mod %>% pluck('fit') %>% mgcv::gam.check() 

```
```{r}
gam_mod %>% pluck('fit') %>% summary() 
gam_mod %>% pluck('fit') %>% plot()
gam_mod %>% pluck('fit') %>% plot( all.terms = TRUE, pages = 1)
```

```{r}
#fitting a more simple GAM model by only allowing for a non-linear function if the edf is greater than 3.
gam_mod2 <- fit(gam_spec,
    new_cases ~ s(total_cases) + s(total_deaths) + s(total_deaths) + s(reproduction_rate) + s(people_vaccinated_per_hundred) + people_fully_vaccinated_per_hundred + tests_per_case + icu_patients + s(positive_rate) + s(total_boosters_per_hundred),
    data = covid
)

gam_mod2 %>% pluck('fit') %>% summary() 
gam_mod2 %>% pluck('fit') %>% plot(all.terms = TRUE, pages=1)
``` 


```{r}

lm_spec <-
  linear_reg() %>%
  set_engine(engine = 'lm') %>%
  set_mode('regression')

data_cv8 <- covid %>% 
    vfold_cv(v = 8)

covid_rec_spline <- recipe(new_cases ~ total_cases + total_deaths + reproduction_rate + people_vaccinated_per_hundred + people_fully_vaccinated_per_hundred+ total_boosters_per_hundred+ tests_per_case + icu_patients+ positive_rate, data = covid)
  
spline_rec <- covid_rec_spline %>% 
  step_ns(total_cases, deg_free = 9) %>%
  step_ns(total_deaths, deg_free = 5) %>% 
  step_ns(reproduction_rate, deg_free = 9) %>% 
  step_ns(people_vaccinated_per_hundred, deg_free = 2) %>% 
  step_ns(total_boosters_per_hundred, deg_free = 8) %>%
  step_ns(positive_rate, deg_free = 1)

# Check the pre-processed data
spline_rec %>% prep(covid) %>% juice()

covid_spline_wf <- workflow() %>%
    add_model(lm_spec) %>%
    add_recipe(covid_rec_spline)
  
spline_wf <- workflow() %>% 
  add_recipe(spline_rec) %>%
  add_model(lm_spec) 

fit_resamples(
    covid_spline_wf,
    resamples = data_cv8, # cv folds
    metrics = metric_set(mae,rmse,rsq)                     
) %>% collect_metrics()

```

```{r}
fit_resamples(
    spline_wf,
    resamples = data_cv8, # cv folds
    metrics = metric_set(mae,rmse,rsq)                     
) %>% collect_metrics()
```

```{r}
ns_mod <- spline_wf %>%
  fit(data = covid) 

ns_mod %>%
  tidy()

spline_mod_output <- covid %>%
  bind_cols(predict(ns_mod, new_data = covid)) %>%
    mutate(resid = new_cases - .pred)

p4 <- ggplot(spline_mod_output, aes(x = total_cases, y = resid)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") +
    theme_classic()
p5 <- ggplot(spline_mod_output, aes(x = total_deaths, y = resid)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") +
    theme_classic()
p6 <- ggplot(spline_mod_output, aes(x = reproduction_rate, y = resid)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") +
    theme_classic()
p7 <- ggplot(spline_mod_output, aes(x = people_vaccinated_per_hundred, y = resid)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") +
    theme_classic()
p8 <- ggplot(spline_mod_output, aes(x = total_boosters_per_hundred, y = resid)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") +
    theme_classic()
p9 <- ggplot(spline_mod_output, aes(x = positive_rate, y = resid)) +
    geom_point() +
    geom_smooth() +
    geom_hline(yintercept = 0, color = "red") +
    theme_classic()
p4
p5
p6
p7
p8
p9
```

```{r, fig.width = 4,fig.height = 1.5}
### RELABEL PLOTS to include lasso and splines for distinction
grid.arrange(pl1, p4, nrow = 1, ncol = 2)
grid.arrange(pl2, p5, nrow = 1, ncol = 2)
grid.arrange(pl3, p6, nrow = 1, ncol = 2)
grid.arrange(pl4, p7, nrow = 1, ncol = 2)
grid.arrange(pl6, p8, nrow = 1, ncol = 2)
grid.arrange(pl9, p9, nrow = 1, ncol = 2)
```

* Compare insights from variable importance analyses here and the corresponding results from the Investigation 1. Now after having accounted for nonlinearity, have the most relevant predictors changed?

--


* Do you gain any insights from the GAM output plots (easily obtained from fitting smoothing splines) for each  predictor?

The GAM output plots reveal a handful of non-linear variables, especially total vaccinations per hundred, and total boosters per hundred. This demonstrates the value of fitting models allowing for non-linearity, as forcing linearity upon these would have resulted in a poor estimate. 
  

* Compare model performance between your GAM models that the models that assuming linearity.
How does test performance of the GAMs compare to other models you explored?

Across all metrics, our GAM models perform somewhat better than those which assume non-linearity. We can see this in the lower MAE, lower RMSE, and slightly higher r-squared.

3. **Summarize investigations**
    - Decide on an overall best model based on your investigations so far. To do this, make clear your analysis goals. Predictive accuracy? Interpretability? A combination of both?

Our analysis goal is to predict the quantity of new COVID cases. Given the universal salience of this subject, it is important to have both highly interpretable results as well as a quality predictive model. To attempt to explain the rate of new COVID cases, we have fit a series of least-squares linear regression models, as well as a series of models subjected to LASSO linear regression. Of our fit models, we find that the best-fit LASSO model has the lowest mean absolute error (MAE), therefore producing the most accurate prediction of our fit models. However, even this model results in a sizable MAE. To correct this, we must consider the non linearity of some input variables, and undertake steps to model them using polynomial regression.  


<br>

4. **Societal impact**
    - Are there any harms that may come from your analyses and/or how the data were collected?
    
    These analyses focus on the COVID-19 pandemic which has resulted in widespread illness and death. The data reflects the very real tragedies of people around the world, and as such should be considered with gravity. These data have been collected in an anonymous way to respect the privacy of those they represent, and can be used in the important effort to understand better the spread of the disease. We do not anticipate incurring any societal harm through the use and analysis of this data. 
    
    - What cautions do you want to keep in mind when communicating your work?
    
It is important to remember that the reporting of these data has been done individually by each country, and therefore is incomplete and prone to some inconsistencies. Additionally, it is important to remember the people behind these data, both for context and in recognition of the lives lost. 




<br><br><br>
