```{r hw2_setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

# Homework 4

## Required Analysis

For this homework,

1. **Specify the research question for a classification task.**

Can we predict the continent the observation comes from?

2. **Try to implement at least 2 different classification methods to answer your research question.**

1. Logistic regression
2. Decision tree

3. **Reflect on the information gained from these two methods and how you might justify this method to others.**

#### Your Work {-}

```{r}
# library statements 
library(dplyr)
library(readr)
library(broom)
library(ggplot2)
library(tidymodels)
library(lubridate)
library(gridExtra)
library(probably)
tidymodels_prefer() # Resolves conflicts, prefers tidymodel functions

# read in data
covid <- read_csv("owid-covid-data.csv")
```

```{r}
# data cleaning
covid <- covid %>%
  select(total_cases, new_cases, date, total_deaths, reproduction_rate, people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred, total_boosters_per_hundred, tests_per_case, icu_patients, positive_rate, life_expectancy, population_density, median_age, gdp_per_capita, continent) %>%
  filter(date >= as.Date("2021-08-13"))%>%
  na.omit(covid)
```

- We're considering COVID cases occurring after the first recorded booster vaccines (13th August, 2021) to best capture today's landscape.  
- This is when vaccinations started to become widespread (at least in the United States).
- In addition to our existing variables, we also added the following to our model to predict the continent:
    - `life_expectancy`
    - `population_density`
    - `median_age`
    - `gdp_per_capita`

```{r}
# creation of cv folds
set.seed(123)
covid_cv10 <- vfold_cv(covid, v = 10)
```

## Classification - Methods

### LASSO Logistic Regression 

We wil fit a LASSO logistic regression model for the `continent` outcome, and allow all possible predictors to be considered.

We initially try a sequence of 100 $\lambda$'s from 1 to 10, which we then adjust based on the plot of test AUC versus $\lambda$.

Since we can only perform binominal logistic regression, we had to create a new variable, `NorthAmerica_or_Not`, categorizing the data based on whether its `continent` was "North America" or not. This new variable will be the outcome we try to predict using this method, setting the reference level as not North America.

```{r}
# Create new binary variable
covid <-covid %>% 
  mutate(NorthAmerica_or_Not = ifelse(continent == 'North America', 'North_America', 'Not_North_America'))


# Set reference level (to the outcome you are NOT interested in)
covid <- covid %>%
  mutate(NorthAmerica_or_Not = relevel(factor(NorthAmerica_or_Not), ref='Not_North_America')) #set reference level

# Logistic LASSO Regression Model Spec
logistic_lasso_spec_tune <- logistic_reg() %>%
    set_engine('glmnet') %>%
    set_args(mixture = 1, penalty = tune()) %>%
    set_mode('classification')

# Recipe
logistic_rec <- recipe(NorthAmerica_or_Not ~ total_cases + new_cases + total_deaths + reproduction_rate + people_vaccinated_per_hundred + people_fully_vaccinated_per_hundred + total_boosters_per_hundred + tests_per_case + icu_patients + positive_rate + life_expectancy + population_density + median_age + gdp_per_capita, data = covid) %>%
    step_normalize(all_numeric_predictors()) %>% 
    step_dummy(all_nominal_predictors())

# Workflow (Recipe + Model)
log_lasso_wf <- workflow() %>% 
    add_recipe(logistic_rec) %>%
    add_model(logistic_lasso_spec_tune) 

# Tune Model (trying a variety of values of Lambda penalty)
penalty_grid <- grid_regular(
  penalty(range = c(-5, 1)), #log10 transformed  (kept moving min down from 0)
  levels = 100)

tune_output <- tune_grid( 
  log_lasso_wf, # workflow
  resamples = covid_cv10, # cv folds
  metrics = metric_set(roc_auc,accuracy),
  control = control_resamples(save_pred = TRUE, event_level = 'second'),
  grid = penalty_grid # penalty grid defined above
)

tune_output
# Visualize Model Evaluation Metrics from Tuning
#autoplot(tune_output) + theme_classic()
```

#### Inspecting the model 

We will choose the final model whose CV AUC is within one standard error of the overall best metric. 

```{r}
# Select Penalty
best_se_penalty <- select_by_one_std_err(tune_output, metric = 'roc_auc', desc(penalty)) # choose penalty value based on the largest penalty within 1 se of the highest CV roc_auc
best_se_penalty

# Fit Final Model
final_fit_se <- finalize_workflow(log_lasso_wf, best_se_penalty) %>% # incorporates penalty value to workflow 
    fit(data = covid)

final_fit_se %>% tidy() %>%
  filter(estimate == 0)
```
With the set of penalty values that we used, the following variables had their coefficients set to 0:

It turns out that these variables were not important in predicting the continent, after accounting for the other features in the model.

Here we will look at how long a variable stayed in the model as a measure of variable importance:

```{r}
glmnet_output <- final_fit_se %>% extract_fit_engine()
    
# Create a boolean matrix (predictors x lambdas) of variable exclusion
bool_predictor_exclude <- glmnet_output$beta==0

# Loop over each variable
var_imp <- sapply(seq_len(nrow(bool_predictor_exclude)), function(row) {
    # Extract coefficient path (sorted from highest to lowest lambda)
    this_coeff_path <- bool_predictor_exclude[row,]
    # Compute and return the # of lambdas until this variable is out forever
    ncol(bool_predictor_exclude) - which.min(this_coeff_path) + 1
})

# Create a dataset of this information and sort
var_imp_data <- tibble(
    var_name = rownames(bool_predictor_exclude),
    var_imp = var_imp
)
var_imp_data %>% arrange(desc(var_imp))
```

Based on the output above, the variables deemed the most important in predicting the continent based on how long they persisted in the LASSO algorithm as the penalty increased.

#### Interpreting evaluation metrics

We inspect the overall CV results for the "best" $\lambda$, and compute the no-information (NIR). 

```{r}
# CV results for "best lambda"
tune_output %>%
    collect_metrics() %>%
    filter(penalty == best_se_penalty %>% pull(penalty))
           
# Count up the continent outcomes
covid %>% count(continent) # Name of the outcome variable goes inside count()

# Compute the NIR
2788/(2788+1813)
```

The AUC for our model is ___. We want it to be as close to 1 as possible (an absolutely perfect classifier of continent). 

#### Choosing a threshold and final model

After using LASSO and balancing bias and variance, we use the final model to make predictions: 

```{r}
# Soft Predictions on Training Data
final_output <- final_fit_se %>% predict(new_data = covid, type='prob') %>% bind_cols(covid)

# FIX THIS
final_output %>%
  ggplot(aes(x = covid, y = .pred_spam)) +
  geom_boxplot()

# ROC curve of sensitivity and specificity
final_output %>%
    roc_curve(covid,.pred_spam,event_level = 'second') %>%
    autoplot()
```

This ROC curve is ___. 

Sensitivity is a measure of how often non-North-American continents are correctly predicted to be not North-America by our model. 

Specificity is the measure of how North-American continents are correctly predicted as North-American by our model.

In this data context, we care more about the true positive rate - sensitivity. A possible application of this is if a user of this data wants to predict a COVID outbreak by continent, in which case we want to correctly predict a continent.

Using the J index

```{r}
# thresholds in terms of reference level
threshold_output <- final_output %>%
    threshold_perf(truth = covid, estimate = .pred_not_spam, thresholds = seq(0,1,by=.01)) 

# J-index v. threshold for not_spam
threshold_output %>%
    filter(.metric == 'j_index') %>%
    ggplot(aes(x = .threshold, y = .estimate)) +
    geom_line() +
    labs(y = 'J-index', x = 'threshold') +
    theme_classic()
```

```{r}
threshold_output %>%
    filter(.metric == 'j_index') %>%
    arrange(desc(.estimate))
```

```{r}
# Distance v. threshold for not_spam

threshold_output %>%
    filter(.metric == 'distance') %>%
    ggplot(aes(x = .threshold, y = .estimate)) +
    geom_line() +
    labs(y = 'Distance', x = 'threshold') +
    theme_classic()
```

```{r}
threshold_output %>%
    filter(.metric == 'distance') %>%
    arrange(.estimate)
```

```{r}
log_metrics <- metric_set(accuracy,sens,yardstick::spec)

final_output %>%
    mutate(.pred_class = make_two_class_pred(.pred_not_spam, levels(spam), threshold = .64)) %>%
    log_metrics(truth = spam, estimate = .pred_class, event_level = 'second')
```



1. **Indicate at least 2 different methods used to answer your classification research question.**
2. **Describe what you did to evaluate the models explored.**
3. **Indicate how you estimated quantitative evaluation metrics.**
4. **Describe the goals / purpose of the methods used in the overall context of your research investigations.**

## Classification - Results

1. **Summarize your final model and justify your model choice (see below for ways to justify your choice)**.
2. **Compare the different classification models tried in light of evaluation metrics, variable importance, and data context.**
3. **Display evaluation metrics for different models in a clean, organized way. This display should include both the estimated metric as well as its standard deviation. (This won’t be available from OOB error estimation. If using OOB, don’t worry about reporting the SD.)**
4. **Broadly summarize conclusions from looking at these evaluation metrics and their measures of uncertainty.**

## Classification - Conclusions

1. **Interpret evaluation metric(s) for the final model in context. Does the model show an acceptable amount of error?**
2. **If using OOB error estimation, display the test (OOB) confusion matrix, and use it to interpret the strengths and weaknesses of the final model.**
3. **Summarization should show evidence of acknowledging the data context in thinking about the sensibility of these results.**
